Load Shifting Versus Manual Frequency Reserve: \\ Which One is More Appealing to Flexible Loads in Denmark?

### 2024-03-07 tex/paper_flex_v2.tex

\begin{document}
\title{
    %A Pioneering Market Framework for Promoting Demand-side Flexibility in Denmark 
    %A New Ecosystem for Demand-side Flexibility Aggregators in Denmark 
    Load Shifting Versus Manual Frequency Reserve: \\ Which One is More Appealing to Flexible Loads in Denmark?
}
% Pioneering Market Framework for the Future Power System: the Case of Denmark
%\author{Peter Alexander Vistar Gade$^{\textrm{a}}$, Henrik Bindner, Jalal
%Kazempour}
\author[1,2]{Peter A.V. Gade\footnote{Corresponding author. Tel.: +45 24263865. \\ Email addresses: pega@dtu.dk (P.A.V. Gade), Trygve.Skjotskift@ibm.com (T. Skjøtskift), chazi@dtu.dk (C. Ziras), hwbi@dtu.dk (H.W. Bindner), jalal@dtu.dk (J. Kazempour).}}
\author[2]{Trygve Skjøtskift}
\author[1]{Charalampos Ziras}
\author[1]{Henrik W. Bindner}
\author[1]{Jalal Kazempour}
\affil[1]{Department of Wind and Energy Systems, Technical University of Denmark, Kgs. Lyngby, Denmark}
\affil[2]{IBM Client Innovation Center, Copenhagen, Denmark}
%\setcounter{Maxaffil}{0}
\renewcommand\Affilfont{\itshape\small}

@@

This paper investigates %how  thermostatically controlled loads can deliver flexibility 
demand-side flexibility provision in two  forms of manual Frequency Restoration Reserve (mFRR) services and load shifting, and explores which one is financially more appealing to flexible loads in Denmark. While mFRR is an ancillary service required for the system and being bought by the system operator, load shifting is an individual act of the flexible load in response to the variation of hourly electricity prices.  %We discuss advantages and disadvantages of the two and their appeal from a monetary point of view.
Without loss of generalization, we consider a supermarket freezer  as a representative flexible load, and develop a grey-box model  describing its temperature dynamics using real data from a supermarket in Denmark. Taking into account price and activation uncertainties, a stochastic mixed-integer linear program is formulated to maximize the flexibility value from the freezer. %For practical reasons, we propose a linear policy to determine regulating power bids.
%, and then linearize the mFRR activation conditions. 
%For computational ease, we develop a decomposition technique, splitting the problem to a set of smaller sub-problems, one per scenario. %An out-of-sample evaluation was done on unseen Danish market price data for 2022. 
%
%We observed that load shifting was more profitable, but had a greater impact on the air and food temperatures in the freezer as opposed to mFRR that depends on the system state and bid price for activation.
%
%solution strategies for mFRR are presented: one with a simple policy and the other one with a dynamically updated policy. The proposed model accurately captures the timeline of decisions for bidding in manual frequency reserves.
%by using McCormick relaxation.
Examined on an ex-post simulation based on  Danish spot and balancing market prices in 2022, load shifting shows to be more profitable than mFRR provision, but is also more consequential for temperature deviations in the freezer. This indicates the need for regulatory measures by the Danish  system operator to make the provision of ancillary services more attractive to  flexible loads.

@@

Power consumers are looking to reduce costs by any means due to sky-rocketing energy and gas prices. In Denmark, supermarkets are especially exposed due to their high energy consumption, and they are actively looking into initiatives to reduce their electricity bill. One of these initiatives is the application of demand-side flexibility \cite{Iacopo,towards}, whereby a supermarket can shift consumption in time when spot (day-ahead) prices are comparatively lower or when the power system needs up-regulation services. There are thousands of supermarkets in Denmark willing to provide flexibility, and their freezers are especially suited for this as they constitute a big share of their energy consumption.

@@

\subsection{Dilemma: Load Shifting Versus mFRR}
%
For power consumers who are actively looking into cost saving solutions, it is important to estimate the monetary benefit of flexibility. In Nord Pool, spot prices are announced at 2pm for the next day, so supermarkets who purchase power through a retailer at  spot prices, have a chance to shift their consumption plan for the next day, accordingly. From their perspective, it is natural to first look into load shifting as the most obvious and straightforward way to utilize flexibility. Here, load shifting simply refers to shifting consumption to another point in time, typically when spot prices are lower \cite{shift}.

@@

\subsection{Research questions and literature review}
%
Using real data from Denmark, this work aims to investigate and answer the following three research questions: (\textit{i}) how can the flexibility of a supermarket freezer as a flexible load be characterized?
(\textit{ii}) how much monetary benefit can this flexible load earn by the mFRR provision opposed to load shifting? and finally (\textit{iii})
what are the pros and cons of participating in the Danish mFRR market versus load shifting? Unlike the second and third questions, the first one has been addressed extensively in the literature. Therefore, the following literature review focuses on the first question, where we also differentiate our work from the current literature. To the best of our knowledge, there is no  work in the existing literature addressing the second and third questions.

@@

In \cite{schaperow2019simulation}, \cite{chanpiwat2020using}, \cite{moglen2020optimal}, and \cite{moglen2020optimal}, residential air condition units are modeled using up- and down-regulation blocks characterizing flexibility. The blocks are obtained from grey-box models of households \cite{siemann2013performance}. The authors show how such a block formulation can be solved using exact and stochastic dynamic programming in the context of peak shaving demand for a utility in the US. The demand-side flexibility then functions as a hedge towards extreme electricity prices. Although such a use case has not been relevant in Denmark yet, the estimation of flexibility blocks is a novel idea as it avoids having to explicitly integrate a physical model into an optimization problem. However, a tradeoff is that the curse of dimensionality quickly makes dynamic programming computationally intractable for a large portfolio of heterogeneous demand-side assets. Furthermore, it might be difficult to assure the Markov property \cite{MarkovProperty} when reformulating the problem even slightly. Flexibility blocks are also used in \cite{bobo2018offering} for an offering strategy. Here, the flexibility blocks are derived from measurements of residential appliances in the Ecogrid 2.0 project \cite{ecogrid}.

@@

@@

There are several ways to monetize flexibility from a Thermostatically Controlled Load (TCL). In this section, we focus on mFRR and load shifting. First, we describe how to mathematically model a TCL as a flexible resource. Second, we describe how to monetize the flexibility from TCLs for mFRR and load shifting, and provide the objective functions in both cases. For mFRR, the objective function includes all costs and revenues for the BRP, while for load shifting, the objective function only includes the flexible demand's perspective. This approach explicitly shows the situation where flexible demand is activated without including the BRP, as this is more realistic.

@@

\subsection{mFRR}\label{sec:mFRR}
%
Fig. \ref{fig:timeline_mfrr} shows the timeline of the mFRR market in Denmark. There is a market for up-regulation only. One should note  that BRPs can choose \textit{not} to bid in the reserve market and  only bid in the real time for up- and down-regulation. We consider a case wherein the flexible demand delivers reservation through a BRP, since the payment is received for both reservation and activation. All prices considered are for the bidding zone DK2 (eastern Denmark).

\begin{figure}[!t]
    \centering
    \includestandalone[width=\columnwidth]{figures/timeline_mfrr_tikz}
    \caption{Timeline of the Danish mFRR market. Symbols $\bm{p}$ denote power (in MW), whereas symbols  $\bm{\lambda}$ refer to price (in DKK/MW).}
    \label{fig:timeline_mfrr}
\end{figure}

According to Fig. \ref{fig:timeline_mfrr}, at 9:30am of day $\rm{D}$-1, the BRP submits reserve capacity bids $p_{h}^{\rm{r},\uparrow}$ $\forall{h}$ to the mFRR market for the next day $\rm{D}$. If accepted, she is paid at the reservation price $\lambda_{h}^{\rm{r}}$. Therefore, she earns $\lambda_{h}^{\rm{r}}p_{h}^{\rm{r},\uparrow}$. This happens \textit{before} the day-ahead (spot) market clearing at noon for which the BRP buys energy for her expected demand $P_{h}^{\rm{Base}}$ at the spot price $\lambda_{h}^{\rm{s}}$. After that, at 5pm, a regulating power bid, $\lambda_{h}^{\rm{bid}}$, must be submitted by the BRP for each hour of day $\rm{D}$, where $p_{h}^{\rm{r},\uparrow} > 0$ \cite{energinet:Systemydelser}. In hour $h$ of day $\rm{D}$, the reserves are activated if the following three conditions hold: (\textit{i}) there is a reservation accepted, i.e., $p_{h}^{\rm{r},\uparrow} > 0$, (\textit{ii}) the balancing price is higher than the spot price, i.e., $\lambda_{h}^{\rm{b}} > \lambda_{h}^{\rm{s}}$, and the regulating power bid is lower than or equal to the balancing price minus the spot price, i.e., $\lambda_{h}^{\rm{bid}} \leq  \lambda_{h}^{\rm{b}} - \lambda_{h}^{\rm{s}}$. If all these three conditions are met, the BRP is paid at the balancing price $\lambda_{h}^{\rm{b}}$ times her actual up-regulation $p_{h}^{\rm{b},\uparrow}$. The BRP may  incur a cost due to any subsequent rebound $p_{h}^{\rm{b},\downarrow}$. Furthermore, the BRP should pay at the penalty price $\lambda^{\rm{p}}$ for $s_{h} = \text{max}\{0, p_{h}^{\rm{r},\uparrow} - p_{h}^{\rm{b},\uparrow}$\}, i.e., if she fails delivering her promised reserve. In reality, $p_{h}^{\rm{b},\uparrow}$ is determined by the TSO, but for simplicity, we assume a bid is always activated in its entirety.
Accordingly, the objective function of the BRP maximizing the flexibility value over a day (via mFRR provision) is
%
\begin{align}\label{eq:mFRRObjective}
    % & \underbrace{C(\rm{cost})}_{\rm{mFRR}} = - \underbrace{\sum_{h=1}^{24}
     & - \underbrace{\sum_{h=1}^{24} \lambda^{\rm{s}}_{h}P^{\rm{Base}}_{h}}_{\textrm{Energy cost}} + \underbrace{\sum_{h=1}^{24}\lambda_{h}^{\rm{r}} p^{\rm{r}, \uparrow}_{h}}_{\textrm{Reservation payment}}  \notag \\ & \quad \quad + \underbrace{\sum_{h=1}^{24}  \lambda_{h}^{\rm{b}} p^{\rm{b},\uparrow}_{h}}_{\textrm{Activation payment}} - \underbrace{\sum_{h=1}^{24}  \lambda_{h}^{\rm{b}} p^{\rm{b},\downarrow}_{h}}_{\textrm{Rebound cost}} - \underbrace{ \sum_{h=1}^{24}  \lambda^{\rm{p}}s_{h}}_{\textrm{Penalty cost}}.
\end{align}

Note that the first term of \eqref{eq:mFRRObjective}, i.e., the energy cost, contains $\lambda^{\rm{s}}_{h}$ and $P^{\rm{Base}}_{h}$, which are both parameters, and therefore, can be omitted from the objective function. However, we include it here for the completeness.

\subsection{Load shifting}
%
Another option for utilizing flexibility is to shift the load to a different time according to the spot market prices which are known already 12-36 hours in advance. Then, it is simply a matter of consuming in low-price hours.
%
For a TCL, there are additional constraints to how energy can be shifted due to the rebound effect. First, there can be temperature constraints which will result in less energy being shifted. Second, the rebound must happen immediately after reducing power consumption. Otherwise, the temperature deviation becomes too big for too long.
The cost saving from load shifting is directly proportional to the volume and price difference between the baseline load and shifted load as given by
%
\begin{equation}\label{eq:load_shifting_savings}
    \sum_{h=1}^{24} \lambda^{\rm{s}}_{h} \big(p^{\rm{Base}}_{h} -  p_{h}\big),
\end{equation}
%
where $p_{h}$ is the power profile when the load is shifted, whereas $p^{\rm{Base}}_{h}$ is the baseline in the day-ahead stage.

Since the load shifting action only occurs \textit{after} the day-ahead market clearing (cf. Fig. \ref{fig:timeline_mfrr}), the BRP has already bought $\lambda^{\rm{s}}_{h} p^{\rm{Base}}_{h}$ and any deviation results in an imbalance for the BRP. In this work, we look at the case where the flexible demand acts selfishly and excludes the BRP from her load shifting action. Therefore, the objective function for the flexible demand simply reduces to $\sum_{h=1}^{24} \lambda_{h}^{\rm{s}} p_{h}$.

@@

\subsection{Time sequence for decision making}
Fig. \ref{fig:timeline_mfrr_variables} shows the stages for making decisions in the mFRR market. In the first stage in day $\rm{D}$-1, the BRP makes a reservation bid decision $p_{h}^{\rm{r},\uparrow}$ for every hour $h$ of day $\rm{D}$,  while being uncertain about input parameters in the next stages, including day-ahead market prices $ \lambda_{h}^{\rm{s}}$  and balancing market prices $ \lambda_{h}^{\rm{b}}$. We assume the BRP is not uncertain about mFRR market prices $ \lambda_{h}^{\rm{r}}$.

For simplicity and avoid the need for developing a multi-stage stochastic program, we merge the second, third, and fourth stages in Fig. \ref{fig:timeline_mfrr_variables} as one, and call it the second stage. By this, $p_{h}^{\rm{r},\uparrow}$ is the first-stage variable, whereas the second stage variables, indexed by scenario $\omega$, include
the regulating power bid $\lambda_{h,\omega}^{\rm{bid}}$ and the set of real-time variables $\Gamma_{h,\omega} = \{ p_{h,\omega}, p_{h,\omega}^{\rm{b},\uparrow}$, $p_{h,\omega}^{\rm{b},\downarrow}$, $s_{h,\omega}$, $T_{h,\omega}^{\rm{c}}$, $T_{h,\omega}^{\rm{f}}$, $T_{h,\omega}^{\rm{c,\text{Base}}}$, $T_{h,\omega}^{\rm{f,\text{Base}}}$, $\phi_{h,\omega}$, $g_{h,\omega}, u^{\uparrow}_{h,\omega}, u^{\downarrow}_{h,\omega}, y^{\uparrow}_{h,\omega}, y^{\downarrow}_{h,\omega}, z^{\uparrow}_{h,\omega}, z^{\downarrow}_{h,\omega} \}$. This set contains the real-time power activated, as well as auxiliary variables for identifying up- and down-regulation,  temperature dynamics, and when to deliver up-regulation according to the bid and prices. See the Appendix for a detailed description. Note that down-regulation refers to the rebound effect of the freezer.
% \begin{figure}[!t]\label{fig:timeline_mfrr_variables}
%     \centering
%     \includegraphics[width=\columnwidth]{figures/timeline_mfrr_variables.png}
%     \caption{Variables related to mFRR up-regulation decisions. The asterisk indicates the first-stage decision.}
% \end{figure}

\begin{figure}[b]
    \centering
    \includestandalone[width=\columnwidth]{figures/timeline_mfrr_variables_tikz}
    \caption{Variables related to the mFRR  bidding decisions.}
    \label{fig:timeline_mfrr_variables}
\end{figure}

\subsection{Scenario generation}\label{sec:scenario_generation}
To make decisions on reservation capacity $p_{h}^{\rm{r},\uparrow}$ and regulating power bid $\lambda_{h,\omega}^{\rm{bid}}$, we generate a set of scenarios $\omega \in \Omega$ for spot and balancing price data., i.e., $ \lambda_{h,\omega}^{\rm{s}}$ and $ \lambda_{h,\omega}^{\rm{b}}$. Each scenario contains spot and balancing price data for the entire day in question. The number of scenarios is $|\Omega|$, and we assume all scenarios are equiprobable. We refer to them as \textit{in-sample} scenarios. We use two different strategies for in-sample scenario generation: (\textit{i}) considering historical spot and balancing prices in DK2 in 2021.
%, with different cases where the number of scenarios $|\Omega|$ is 1, 5, 10, 20, 30, 40, 50, 100, and 250. 
(\textit{ii}) considering prices of the most recent five days (lookback strategy). In both scenario generation strategies, balancing prices $\lambda_{h,\omega}^{\rm{b}}$ are sampled in the following way: First, an integer $v$ is sampled uniformly from $\{0, \ldots, 24\}$ which represents the total number of up-regulation hours in a day. We then sample spot and balancing price differentials from a single day within the set of all days where up-regulation happened $v$ times. This constitutes one scenario and is repeated $|\Omega|$ times. In this way, days where up-regulation happened are essentially up-sampled, and the model learns more when up-regulation happens than otherwise. The in-sample results are systematically compared against the same set of unseen \textit{Out-of-Sample} (OOS) scenarios, which are DK2 prices for 2022.

@@

In the case of load shifting, it is worth mentioning that the settlement cost of the BRP is ignored --- this reflects the self-interested flexible loads.  While the flexible load reduces her cost by load shifting, the corresponding BRP may incur a settlement cost due to the resulting imbalance between the true consumption and the day-ahead schedule. In practice, the BRP would have to pay for such an imbalance or at least buy/sell energy in the intra-day market to adjust her day-ahead schedule and consider potential load shifting actions by flexible loads. However, loads are not necessarily aware of those costs for the BRP. Hence, they may have a strong incentive to exploit their flexibility for load shifting. Some larger flexible consumers, such as industrial and commercial loads, might not be exposed fully to spot prices, and for those consumers, load shifting could be less profitable. However, they may still have an incentive to change their deal with the BRP to get a full exposure.

In the case of mFRR  provision, it is   assumed that the entire revenue from reservation and activation go to the flexible load. This neglects the fact that, in practice, the BRP requires a share of the revenue. Furthermore, there might be an aggregator or a technology provider who facilitates the aggregation and communication of the flexibility. In such a case, they would also request a share of the revenue. This can potentially reduce the revenue of flexible loads by the mFRR provision.

>>>>

\begin{document}
\title{
    %A Pioneering Market Framework for Promoting Demand-side Flexibility in Denmark 
    %A New Ecosystem for Demand-side Flexibility Aggregators in Denmark 
    Load Shifting Versus Manual Frequency Reserve: \\ Which One is More Appealing \textcolor{red}{to Thermostatically Controlled Loads in Denmark}?
}
% Pioneering Market Framework for the Future Power System: the Case of Denmark
%\author{Peter Alexander Vistar Gade$^{\textrm{a}}$, Henrik Bindner, Jalal
%Kazempour}
\author[1,2]{Peter A.V. Gade\footnote{Corresponding author. Tel.: +45 24263865. \\ Email addresses: pega@dtu.dk (P.A.V. Gade), Trygve.Skjotskift@ibm.com (T. Skjøtskift), chazi@dtu.dk (C. Ziras), hwbi@dtu.dk (H.W. Bindner), jalal@dtu.dk (J. Kazempour).}}
\author[2]{Trygve Skjøtskift}
\author[1]{Charalampos Ziras}
\author[1]{Henrik W. Bindner}
\author[1]{Jalal Kazempour}
\affil[1]{Department of Wind and Energy Systems, Technical University of Denmark, Kgs. Lyngby, Denmark}
\affil[2]{IBM Client Innovation Center, Copenhagen, Denmark}
%\setcounter{Maxaffil}{0}
\renewcommand\Affilfont{\itshape\small}

@@

This paper investigates %how  thermostatically controlled loads can deliver flexibility 
demand-side flexibility provision in two  forms of manual Frequency Restoration Reserve (mFRR) services and load shifting, and explores which one is financially more appealing to \textcolor{red}{Thermostatically Controlled Loads (TCLs)} in Denmark. While mFRR is an ancillary service required for the system and being bought by the system operator, load shifting is an individual act of the \textcolor{red}{TCL} in response to the variation of hourly electricity prices.  %We discuss advantages and disadvantages of the two and their appeal from a monetary point of view.
Without loss of generalization, we consider a supermarket freezer  as a representative \textcolor{red}{TCL}, and develop a grey-box model  describing its temperature dynamics using real data from a supermarket in Denmark. Taking into account price and activation uncertainties, a stochastic mixed-integer linear program is formulated to maximize the flexibility value from the freezer. %For practical reasons, we propose a linear policy to determine regulating power bids.
%, and then linearize the mFRR activation conditions. 
%For computational ease, we develop a decomposition technique, splitting the problem to a set of smaller sub-problems, one per scenario. %An out-of-sample evaluation was done on unseen Danish market price data for 2022. 
%
%We observed that load shifting was more profitable, but had a greater impact on the air and food temperatures in the freezer as opposed to mFRR that depends on the system state and bid price for activation.
%
%solution strategies for mFRR are presented: one with a simple policy and the other one with a dynamically updated policy. The proposed model accurately captures the timeline of decisions for bidding in manual frequency reserves.
%by using McCormick relaxation.
Examined on an ex-post simulation based on  Danish spot and balancing market prices in 2022, load shifting shows to be more profitable than mFRR provision, but is also more consequential for temperature deviations in the freezer. This indicates the need for regulatory measures by the Danish  system operator to make the provision of ancillary services more attractive to  \textcolor{red}{TCL}s.

@@

Power consumers are looking to reduce costs by any means due to sky-rocketing energy and gas prices. In Denmark, supermarkets are especially exposed due to their high energy consumption, and they are actively looking into initiatives to reduce their electricity bill. One of these initiatives is the application of demand-side flexibility, whereby a supermarket can shift consumption in time when spot (day-ahead) prices are comparatively lower or when the power system needs up-regulation services. There are thousands of supermarkets in Denmark willing to provide flexibility, and their freezers are especially suited for this as they constitute a big share of their energy consumption.

@@

\subsection{Dilemma: Load Shifting Versus mFRR}
%
For power consumers who are actively looking into cost saving solutions, it is important to estimate the monetary benefit of flexibility. In Nord Pool, spot prices are announced at 2pm for the next day, so supermarkets who purchase power through a retailer at  spot prices, have a chance to shift their consumption plan for the next day, accordingly. From their perspective, it is natural to first look into load shifting as the most obvious and straightforward way to utilize flexibility. Here, load shifting simply refers to shifting consumption to another point in time, typically when spot prices are lower.

@@

\subsection{Research questions and literature review}
%
Using real data from Denmark, this work aims to investigate and answer the following three research questions: (\textit{i}) how can the flexibility of a supermarket freezer as a \textcolor{red}{TCL} be characterized?
(\textit{ii}) how much monetary benefit can this \textcolor{red}{TCL} earn by the mFRR provision opposed to load shifting? and finally (\textit{iii})
what are the pros and cons of participating in the Danish mFRR market versus load shifting? Unlike the second and third questions, the first one has been addressed extensively in the literature. Therefore, the following literature review focuses on the first question, where we also differentiate our work from the current literature. To the best of our knowledge, there is no  work in the existing literature addressing the second and third questions.

@@

\textcolor{red}{In \cite{schaperow2019simulation,chanpiwat2020using, moglen2020optimal}}, residential air condition units are modeled using up- and down-regulation blocks characterizing flexibility. The blocks are obtained from grey-box models of households \cite{siemann2013performance}. The authors show how such a block formulation can be solved using exact and stochastic dynamic programming in the context of peak shaving demand for a utility in the US. The demand-side flexibility then functions as a hedge towards extreme electricity prices. Although such a use case has not been relevant in Denmark yet, the estimation of flexibility blocks is a novel idea as it avoids having to explicitly integrate a physical model into an optimization problem. However, a tradeoff is that the curse of dimensionality quickly makes dynamic programming computationally intractable for a large portfolio of heterogeneous demand-side assets. Furthermore, it might be difficult to assure the Markov property \cite{MarkovProperty} when reformulating the problem even slightly. Flexibility blocks are also used in \cite{bobo2018offering} for an offering strategy. Here, the flexibility blocks are derived from measurements of residential appliances in the Ecogrid 2.0 project \cite{ecogrid}.

@@

\textcolor{red}{Similar studies exist in the literature albeit in different settings. For example, \cite{iria2018trading} uses a two-stage stochastic optimization for prosumers to bid into tertiary reserves, i.e., mFRR. A similar approach is taken in \cite{la2021mixed} although with a focus on scalability to thousands of assets and distribution of flexibility amongst prosumers. In \cite{nitsch2021economic}, the authors show how batteries can be used to provide secondary reserves. TCLs have also been investigated for provision of primary reserves, i.e., frequency reserves, in the Nordics \cite{paridari2020flexibility}, in microgrids \cite{mendieta2020primary}, and in the Australian power grid \cite{attarha2020network}}.

@@

There are several ways to monetize flexibility from a \textcolor{red}{TCL}. In this section, we focus on mFRR and load shifting. First, we describe how to mathematically model a TCL as a flexible resource. Second, we describe how to monetize the flexibility from TCLs for mFRR and load shifting, and provide the objective functions in both cases. For mFRR, the objective function includes all costs and revenues for the BRP, while for load shifting, the objective function only includes the flexible demand's perspective. This approach explicitly shows the situation where flexible demand is activated without including the BRP, as this is more realistic.

@@

\subsection{mFRR}\label{sec:mFRR}
%
Fig. \ref{fig:timeline_mfrr_variables} shows the timeline of the mFRR market in Denmark. There is a market for up-regulation only. One should note  that BRPs can choose \textit{not} to bid in the reserve market and  only bid in the real time for up- and down-regulation. We consider a case wherein the flexible demand delivers reservation through a BRP, since the payment is received for both reservation and activation. All prices considered are for the bidding zone DK2 (eastern Denmark).

% \begin{figure}[!t]
%     \centering
%     \includestandalone[width=\columnwidth]{figures/timeline_mfrr_tikz}
%     \caption{Timeline of the Danish mFRR market. Symbols $\bm{p}$ denote power (in MW), whereas symbols  $\bm{\lambda}$ refer to price (in DKK/MW).}
%     \label{fig:timeline_mfrr}
% \end{figure}

\begin{figure}[b]
    \centering
    \includestandalone[width=\columnwidth]{figures/timeline_mfrr_variables_tikz}
    \caption{\textcolor{red}{Variables related to the timeline of mFRR bidding decisions and load shifting.}}
    \label{fig:timeline_mfrr_variables}
\end{figure}

According to Fig. \ref{fig:timeline_mfrr_variables}, at 9:30am of day $\rm{D}$-1, the BRP submits reserve capacity bids $p_{h}^{\rm{r},\uparrow}$ $\forall{h}$ to the mFRR market for the next day $\rm{D}$. If accepted, she is paid at the reservation price $\lambda_{h}^{\rm{r}}$. Therefore, she earns $\lambda_{h}^{\rm{r}}p_{h}^{\rm{r},\uparrow}$. This happens \textit{before} the day-ahead (spot) market clearing at noon for which the BRP buys energy for her expected demand $P_{h}^{\rm{Base}}$ at the spot price $\lambda_{h}^{\rm{s}}$, \textcolor{red}{(which is not a decision variable in our model)}. After that, at 5pm, a regulating power bid, $\lambda_{h}^{\rm{bid}}$, must be submitted by the BRP for each hour of day $\rm{D}$, where $p_{h}^{\rm{r},\uparrow} > 0$ \cite{energinet:Systemydelser}. In hour $h$ of day $\rm{D}$, the reserves are activated if the following three conditions hold: (\textit{i}) there is a reservation accepted, i.e., $p_{h}^{\rm{r},\uparrow} > 0$, (\textit{ii}) the balancing price is higher than the spot price, i.e., $\lambda_{h}^{\rm{b}} > \lambda_{h}^{\rm{s}}$, and the regulating power bid is lower than or equal to the balancing price minus the spot price, i.e., $\lambda_{h}^{\rm{bid}} \leq  \lambda_{h}^{\rm{b}} - \lambda_{h}^{\rm{s}}$. If all these three conditions are met, the BRP is paid at the balancing price $\lambda_{h}^{\rm{b}}$ times her actual up-regulation $p_{h}^{\rm{b},\uparrow}$. The BRP may  incur a cost due to any subsequent rebound $p_{h}^{\rm{b},\downarrow}$. Furthermore, the BRP should pay at the penalty price $\lambda^{\rm{p}}$ for $s_{h} = \text{max}\{0, p_{h}^{\rm{r},\uparrow} - p_{h}^{\rm{b},\uparrow}$\}, i.e., if she fails delivering her promised reserve. In reality, $p_{h}^{\rm{b},\uparrow}$ is determined by the TSO, but for simplicity, we assume a bid is always activated in its entirety.
Accordingly, the objective function of the BRP maximizing the flexibility value over a day (via mFRR provision) is
%
\begin{align}\label{eq:mFRRObjective}
    \underbrace{\sum_{h=1}^{24}\lambda_{h}^{\rm{r}} p^{\rm{r}, \uparrow}_{h}}_{\textrm{Reservation payment}} + \underbrace{\sum_{h=1}^{24}  \lambda_{h}^{\rm{b}} p^{\rm{b},\uparrow}_{h}}_{\textrm{Activation payment}} - \underbrace{\sum_{h=1}^{24}  \lambda_{h}^{\rm{b}} p^{\rm{b},\downarrow}_{h}}_{\textrm{Rebound cost}} - \underbrace{ \sum_{h=1}^{24}  \lambda^{\rm{p}}s_{h}}_{\textrm{Penalty cost}}
\end{align}

\textcolor{red}{We note that an additional term containing fixed energy cost is subtracted from \eqref{eq:mFRRObjective}, i.e., $\sum_{h=1}^{24} \lambda^{\rm{s}}_{h}P^{\rm{Base}}_{h}$. This term contains $\lambda^{\rm{s}}_{h}$ and $P^{\rm{Base}}_{h}$, which are both parameters, and therefore, can be omitted from the objective function. However, we include it for cost comparison with load shifting.}

\subsection{Load shifting}
%
Another option for utilizing flexibility is to shift the load to a different time according to the spot market prices which are known already 12-36 hours in advance \textcolor{red}{as shown in Figure \ref{fig:timeline_mfrr_variables}}. We note that in Europe and Denmark, day-ahead prices are directly exposed to consumers with no time-of-use price in between as is often the practice in North America. Then, it is simply a matter of consuming in low-price hours.
%
For a TCL, there are additional constraints to how energy can be shifted due to the rebound effect. First, there can be temperature constraints which will result in less energy being shifted. Second, the rebound must happen immediately after reducing power consumption. Otherwise, the temperature deviation becomes too big for too long.
The cost saving from load shifting is directly proportional to the volume and price difference between the baseline load and shifted load as given by
%
\begin{equation}\label{eq:load_shifting_savings}
    \sum_{h=1}^{24} \lambda^{\rm{s}}_{h} \big(p^{\rm{Base}}_{h} -  p_{h}\big),
\end{equation}
%
where $p_{h}$ is the power profile when the load is shifted, whereas $p^{\rm{Base}}_{h}$ is the baseline in the day-ahead stage.

Since the load shifting action only occurs \textit{after} the day-ahead market clearing (cf. Fig. \ref{fig:timeline_mfrr_variables}), the BRP has already bought $\lambda^{\rm{s}}_{h} p^{\rm{Base}}_{h}$ and any deviation results in an imbalance for the BRP. In this work, we look at the case where the flexible demand acts selfishly and excludes the BRP from her load shifting action. Therefore, the objective function for the flexible demand simply reduces to $\sum_{h=1}^{24} \lambda_{h}^{\rm{s}} p_{h}$.

@@

\subsection{Time sequence for decision making}
Fig. \ref{fig:timeline_mfrr_variables} shows the stages for making decisions in the mFRR market \textcolor{red}{and for load shifting}. In the first stage in day $\rm{D}$-1, the BRP makes a reservation bid decision $p_{h}^{\rm{r},\uparrow}$ for every hour $h$ of day $\rm{D}$,  while being uncertain about input parameters in the next stages, including day-ahead market prices $ \lambda_{h}^{\rm{s}}$  and balancing market prices $ \lambda_{h}^{\rm{b}}$. We assume the BRP is not uncertain about mFRR market prices $ \lambda_{h}^{\rm{r}}$.

For simplicity and \textcolor{red}{to} avoid the need for developing a multi-stage stochastic program, we merge the second, third, and fourth stages in Fig. \ref{fig:timeline_mfrr_variables} as one, and call it the second stage. By this, $p_{h}^{\rm{r},\uparrow}$ is the first-stage variable, whereas the second stage variables, indexed by scenario $\omega$, include
the regulating power bid $\lambda_{h,\omega}^{\rm{bid}}$ and the set of real-time variables $\Gamma_{h,\omega} = \{ p_{h,\omega}, p_{h,\omega}^{\rm{b},\uparrow}$, $p_{h,\omega}^{\rm{b},\downarrow}$, $s_{h,\omega}$, $T_{h,\omega}^{\rm{c}}$, $T_{h,\omega}^{\rm{f}}$, $T_{h,\omega}^{\rm{c,\text{Base}}}$, $T_{h,\omega}^{\rm{f,\text{Base}}}$, $\phi_{h,\omega}$, $g_{h,\omega}, u^{\uparrow}_{h,\omega}, u^{\downarrow}_{h,\omega}, y^{\uparrow}_{h,\omega}, y^{\downarrow}_{h,\omega}, z^{\uparrow}_{h,\omega}, z^{\downarrow}_{h,\omega} \}$. This set contains the real-time power activated, as well as auxiliary variables for identifying up- and down-regulation,  temperature dynamics, and when to deliver up-regulation according to the bid and prices. See the Appendix for a detailed description. Note that down-regulation refers to the rebound effect of the freezer.
% \begin{figure}[!t]\label{fig:timeline_mfrr_variables}
%     \centering
%     \includegraphics[width=\columnwidth]{figures/timeline_mfrr_variables.png}
%     \caption{Variables related to mFRR up-regulation decisions. The asterisk indicates the first-stage decision.}
% \end{figure}

\subsection{Scenario generation}\label{sec:scenario_generation}
To make decisions on reservation capacity $p_{h}^{\rm{r},\uparrow}$ and regulating power bid $\lambda_{h,\omega}^{\rm{bid}}$, we generate a set of scenarios $\omega \in \Omega$ for spot and balancing price data., i.e., $ \lambda_{h,\omega}^{\rm{s}}$ and $ \lambda_{h,\omega}^{\rm{b}}$. Each scenario contains spot and balancing price data for the entire day in question. The number of scenarios is $|\Omega|$, and we assume all scenarios are equiprobable. We refer to them as \textit{in-sample} scenarios. We use two different strategies for in-sample scenario generation: (\textit{i}) considering historical spot and balancing prices in DK2 in 2021. \textcolor{red}{Solution strategy (i) represents the case where an aggregator can rely on a simple, fixed policy without having to run a daily optimization as opposed to the second solution strategy, although it might prove better in case the underlying uncertainty is non-stationary. The first solution strategy uses 50 scenarios which, and we found that increasing the number of scenarios further did not improve the performance. The same was the case for choosing five days in the second solution strategy.}
%, with different cases where the number of scenarios $|\Omega|$ is 1, 5, 10, 20, 30, 40, 50, 100, and 250. 
(\textit{ii}) considering prices of the most recent five days (lookback strategy). In both scenario generation strategies, balancing prices $\lambda_{h,\omega}^{\rm{b}}$ are sampled in the following way: First, an integer $v$ is sampled uniformly from $\{0, \ldots, 24\}$ which represents the total number of up-regulation hours in a day. We then sample spot and balancing price differentials from a single day within the set of all days where up-regulation happened $v$ times. This constitutes one scenario and is repeated $|\Omega|$ times. In this way, days where up-regulation happened are essentially up-sampled, and the model learns more when up-regulation happens than otherwise. The in-sample results are systematically compared against the same set of unseen \textit{Out-of-Sample} (OOS) scenarios, which are DK2 prices for 2022.

@@

In the case of load shifting, it is worth mentioning that the settlement cost of the BRP is ignored --- this reflects the self-interested \textcolor{red}{TCL}s.  While the \textcolor{red}{TCL} reduces her cost by load shifting, the corresponding BRP may incur a settlement cost due to the resulting imbalance between the true consumption and the day-ahead schedule. In practice, the BRP would have to pay for such an imbalance or at least buy/sell energy in the intra-day market to adjust her day-ahead schedule and consider potential load shifting actions by \textcolor{red}{TCL}s. However, loads are not necessarily aware of those costs for the BRP. Hence, they may have a strong incentive to exploit their flexibility for load shifting. Some larger flexible consumers, such as industrial and commercial loads, might not be exposed fully to spot prices, and for those consumers, load shifting could be less profitable. However, they may still have an incentive to change their deal with the BRP to get a full exposure.
